---
title: "1 — Calculate VAFI Frailty Scores"
author: "Pablo Knobel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: united
---

# Summary

This script calculates the Veterans Affairs Frailty Index (VAFI) for all
patients in the Mount Sinai primary care cohort. For each patient × year
combination (2011–2023), it counts how many of the 31 VAFI deficits are
present in the `year_prior` years before the index year, using ICD-10
diagnosis codes and CPT procedure codes as evidence. The output is a
wide-format dataset with one row per patient × year containing the VAFI
score, the deficit count, and binary deficit indicators.

---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(janitor)
library(lubridate)
library(stringr)
library(arrow)
library(tidyr)
library(zoo)
library(ggplot2)
library(fst)
library(pbapply)
library(reshape2)
```

```{r parameters}
# --- Paths ---
wd              <- "C:/Users/knobep01/OneDrive - The Mount Sinai Hospital"
path_patient    <- file.path(wd, "Maayans lab", "Outcome data", "MSDW 2024 data", "Original files", "PATIENT.csv")
path_diagnosis  <- file.path(wd, "Maayans lab", "Outcome data","MSDW 2024 data", "Original files", "DIAGNOSIS.csv")
path_procedures <- file.path(wd, "Maayans lab", "Outcome data","MSDW 2024 data", "Original files", "IMPRESSION NOTES.csv")
path_icd_map  <- here::here("VA-FI Mappings", "icd_code_vafi_mapping.fst")
path_prcd_map <- here::here("VA-FI Mappings", "procedure_code_vafi_mapping.fst")
path_output     <- file.path(wd, "Knobel Lab", "Birds and Brains", "Birds Frailty - NYC", "vafi_results.rds")

# --- Study parameters ---
study_years <- 2011:2023   # Index years for VAFI calculation
year_prior  <- 3L           # Years of lookback window for deficit ascertainment
chunk_size  <- 1e6          # Rows per processing chunk (memory management)

# --- VAFI deficit columns
# Derive deficit list from mapping files (single source of truth)
cols_deficit <- union(
  fst::read.fst(path_icd_map)$Deficit,
  fst::read.fst(path_prcd_map)$Deficit
) %>% sort()

length(cols_deficit)  
```

---

# Helper Functions

The two internal functions below handle the ICD-9/ICD-10 switchover and the
chunked deficit ascertainment. They are defined here so the main `make_vafi()`
call in the next section is self-contained.

```{r functions}
# Filter ICD codes to respect the ICD-9 → ICD-10 switchover on 2015-10-01.
# Working at year resolution: ICD-9 codes are retained only before 2015;
# ICD-10 codes are retained from 2015 onward; CPT codes are always retained.
.filter_icd_cutoff_years <- function(icd_df) {
  switchover_year <- 2015L
  icd_df %>%
    filter(
      (grepl("ICD9",  CodeType) & Date <  switchover_year) |
      (grepl("ICD10", CodeType) & Date >= switchover_year) |
       CodeType == "CPT"
    )
}

# For a chunk of claim rows, join to the index-date grid, keep only rows
# within the lookback window, then map codes to named deficits.
.make_deficit_chunk_years <- function(chunk_size, icd_df,
                                      year_prior, index_date_tb, mapping) {
  n <- nrow(icd_df)
  pblapply(seq(1, n, by = chunk_size), function(i_lo) {
    i_hi <- min(i_lo + chunk_size - 1L, n)
    icd_df[i_lo:i_hi, ] %>%
      inner_join(
        index_date_tb %>% select(PatientICN, IndexDate),
        by = "PatientICN"
      ) %>%
      mutate(
        Date        = as.numeric(Date),
        IndexDate   = as.numeric(IndexDate),
        date_to_idx = IndexDate - Date
      ) %>%
      filter(IndexDate >= Date, date_to_idx < year_prior) %>%
      select(PatientICN, IndexDate, Code, CodeType) %>%
      distinct() %>%
      inner_join(mapping, by = c("Code", "CodeType")) %>%
      select(PatientICN, IndexDate, Deficit) %>%
      distinct()
  }) %>%
    bind_rows()
}

# Main VAFI function.
# Returns a wide-format data frame: one row per PatientICN × IndexDate,
# with columns vafi (score), one binary column per deficit, and a sentinel
# 0 row for patient-years with no recorded deficits.
make_vafi <- function(icds, prcd, index_date_tb,
                      index_date_col = "year",
                      year_prior     = 3L,
                      chunk_size     = 1e6) {

  index_date_tb[["IndexDate"]] <- as.numeric(index_date_tb[[index_date_col]])

  icd_mapping  <- fst::read.fst(path_icd_map)
  prcd_mapping <- fst::read.fst(path_prcd_map)

  total_deficit <- length(unique(c(icd_mapping$Deficit, prcd_mapping$Deficit)))

  icds_clean <- icds %>%
    transmute(PatientICN, Code = ICDCode, CodeType = ICDCodeType, Date) %>%
    .filter_icd_cutoff_years()

  prcd_clean <- prcd %>%
    transmute(PatientICN, Code = ProcedureCode, CodeType = ProcedureCodeType, Date) %>%
    .filter_icd_cutoff_years()

  cat("Ascertaining deficits from diagnosis codes...\n")
  icd_deficits <- .make_deficit_chunk_years(
    chunk_size, icds_clean, year_prior, index_date_tb, icd_mapping
  )

  cat("Ascertaining deficits from procedure codes...\n")
  prcd_deficits <- .make_deficit_chunk_years(
    chunk_size, prcd_clean, year_prior, index_date_tb, prcd_mapping
  )

  cat("Aggregating deficits and computing VAFI...\n")
  dfct <- list(icd_deficits, prcd_deficits) %>%
    bind_rows() %>%
    distinct() %>%
    group_by(PatientICN, IndexDate) %>%
    mutate(
      num_deficit = n(),
      vafi        = num_deficit / total_deficit,
      Value       = 1L
    ) %>%
    reshape2::dcast(PatientICN + IndexDate + vafi ~ Deficit,
                    value.var = "Value") %>%
    right_join(
      index_date_tb %>% select(PatientICN, IndexDate),
      by = c("PatientICN", "IndexDate")
    ) %>%
    mutate(vafi = if_else(is.na(vafi), 0, vafi))

  dfct[is.na(dfct)] <- 0L
  dfct
}
```

---

# Load Source Data

## Patients

One row per patient. Only the MRN is needed here; it is used to build the
complete patient × year grid.

```{r load-patients}
patient_ids <- read.csv(path_patient) %>%
  clean_names() %>%
  transmute(PatientICN = mrn) %>%
  distinct()

cat("Unique patients:", nrow(patient_ids), "\n")
```

## Index-Date Grid

A complete grid of every patient × index year for the study period.
Each row represents one VAFI measurement occasion.

```{r index-grid}
index_date_tb <- patient_ids %>%
  crossing(Date = study_years)

cat("Index-date grid rows:", nrow(index_date_tb),
    "(", n_distinct(index_date_tb$PatientICN), "patients ×",
    length(study_years), "years)\n")
```

## Diagnosis Codes

```{r load-diagnoses}
icds <- read.csv(path_diagnosis) %>%
  clean_names() %>%
  transmute(
    PatientICN  = mrn,
    ICDCode     = icd10_code,
    ICDCodeType = "ICD10",
    Date        = as.integer(format(as.Date(diagnosis_date), "%Y"))
  )

cat("Diagnosis rows loaded:", nrow(icds), "\n")
```

## Procedure Codes

```{r load-procedures}
prcd <- read.csv(path_procedures) %>%
  clean_names() %>%
  transmute(
    PatientICN        = mrn,
    ProcedureCode     = cpt_code,
    ProcedureCodeType = "CPT",
    Date              = as.integer(format(as.Date(order_dttm), "%Y"))
  )

cat("Procedure rows loaded:", nrow(prcd), "\n")
```

---

# Calculate VAFI

For each patient × index year, deficits are ascertained from claims in the
`year_prior` years before (and including) the index year.

```{r calculate-vafi}
vafi_year <- make_vafi(
  icds           = icds,
  prcd           = prcd,
  index_date_tb  = index_date_tb,
  index_date_col = "Date",
  year_prior     = year_prior,
  chunk_size     = chunk_size
)

cat("VAFI rows:", nrow(vafi_year), "\n")
cat("Unique patients:", n_distinct(vafi_year$PatientICN), "\n")
```

## Add Deficit Count

`deficit_count` is the sum of the 30 binary deficit columns. Patients with
no coded deficits have a deficit count of 0 and a VAFI of 0.

```{r add-deficit-count}

# Add any deficits that were never observed (e.g. DuraMed) as zero columns
missing_deficits <- setdiff(cols_deficit, names(vafi_year))
vafi_year[missing_deficits] <- 0L

vafi_year <- vafi_year %>%
  mutate(
    deficit_count = rowSums(
      across(all_of(cols_deficit)),
      na.rm = TRUE
    )
  )

cat("Deficit count range:", min(vafi_year$deficit_count),
    "-", max(vafi_year$deficit_count), "\n")
cat("Mean deficit count :", round(mean(vafi_year$deficit_count), 2), "\n")
```

---

# Validation

```{r validate}
cat("\n========== VALIDATION ==========\n")

cat("Rows              :", nrow(vafi_year), "\n")
cat("Unique patients   :", n_distinct(vafi_year$PatientICN), "\n")
cat("Years covered     :", min(vafi_year$IndexDate), "-",
                          max(vafi_year$IndexDate), "\n")

cat("\nVAFI distribution:\n")
print(summary(vafi_year$vafi))

cat("\nDeficit count distribution:\n")
print(summary(vafi_year$deficit_count))

cat("\nPatients frail (VAFI > 0.2):",
    sum(vafi_year$vafi > 0.2, na.rm = TRUE),
    sprintf("(%.1f%%)", mean(vafi_year$vafi > 0.2, na.rm = TRUE) * 100), "\n")

stopifnot(
  "VAFI must be between 0 and 1"  = all(vafi_year$vafi >= 0 & vafi_year$vafi <= 1),
  "No negative deficit counts"    = all(vafi_year$deficit_count >= 0),
  "deficit_count consistent with vafi" =
    all(abs(vafi_year$deficit_count / total_deficits - vafi_year$vafi) < 0.01 |
        vafi_year$vafi == 0)
)

cat("All validation checks passed.\n")
```

---

# Plots

## Spaghetti Plot — VAFI Trajectories

A random sample of 1 000 patients illustrates individual-level heterogeneity
in frailty trajectories. The thick line shows the cohort mean.

```{r plot-spaghetti, fig.width = 8, fig.height = 5}
set.seed(42)
sample_ids <- vafi_year %>%
  distinct(PatientICN) %>%
  slice_sample(n = 1000)

vafi_sample <- vafi_year %>%
  semi_join(sample_ids, by = "PatientICN")

fig_trajectories <- ggplot(
  vafi_sample,
  aes(x = IndexDate, y = vafi, group = PatientICN)
) +
  geom_line(alpha = 0.15, colour = "steelblue") +
  stat_summary(
    aes(group = 1),
    fun      = mean,
    geom     = "line",
    linewidth = 1.2,
    colour   = "firebrick"
  ) +
  scale_x_continuous(breaks = study_years) +
  labs(
    title    = "VAFI trajectories — random sample of 1 000 patients",
    subtitle = "Blue = individual; red = cohort mean",
    x        = "Index year",
    y        = "VAFI score"
  ) +
  theme_minimal()

fig_trajectories
```

## Deficit Prevalence Bar Chart

Shows how frequently each deficit was ascertained across all patient-years,
ordered from most to least common.

```{r plot-deficit-prevalence, fig.width = 8, fig.height = 6}
deficit_prevalence <- vafi_year %>%
  select(all_of(cols_deficit)) %>%
  summarise(across(everything(), ~ sum(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(),
               names_to  = "deficit",
               values_to = "count")

fig_deficits <- ggplot(
  deficit_prevalence,
  aes(x = reorder(deficit, count), y = count)
) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Prevalence of each VAFI deficit (total patient-years)",
    x     = "Deficit",
    y     = "Count"
  ) +
  theme_minimal()

fig_deficits
```

---

# Save

```{r save}
saveRDS(vafi_year, path_output)

cat("Saved:", path_output, "\n")
cat("Rows  :", nrow(vafi_year), "\n")
cat("Unique patients:", n_distinct(vafi_year$PatientICN), "\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
