---
title: "2 — Build Analytical Dataset"
author: "Pablo Knobel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: united
---

# Summary

Builds the analytical dataset for the bird biodiversity and frailty analysis.
Starting from the MSDW primary care cohort, the pipeline:

1. Loads patient demographics and derived VAFI frailty scores
2. Links patients to census tracts via geocoded addresses (time-varying)
3. Joins tract-level ACS socioeconomic variables (time-varying, annual)
4. Joins latent Shannon diversity estimates from the eBird GAM (time-varying, annual)
5. Merges encounter records and applies two-consecutive-year trimming,
   with death records protected from premature censoring
6. Removes post-death observations
7. Derives age, binary frailty, and IQR-scaled Shannon exposure
8. Runs validation checks and saves

---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(readr)
library(janitor)
library(lubridate)
library(tidyr)
library(purrr)
library(ggplot2)
library(sf)
library(tidycensus)
```

```{r parameters}
# --- Paths ---
wd <- "C:/Users/knobep01/OneDrive - The Mount Sinai Hospital"

path_patient   <- file.path(wd, "Maayans lab", "Outcome data", "MSDW 2024 data", "Original files", "PATIENT.csv")
path_encounter <- file.path(wd, "Maayans lab", "Outcome data", "MSDW 2024 data", "Original files", "ENCOUNTER.csv")
path_geocodes  <- file.path(wd, "Maayans lab", "Helena",
                            "MSDW PC patients 2011-2023",
                            "generated files", "Shapefiles and goecodes",
                            "Geocodes", "final addresses dates.rds")
path_vafi      <- file.path(wd, "Knobel Lab", "Birds and Brains", "Birds Frailty - NYC", "vafi_results.rds")
path_shannon   <- file.path(wd, "Non PHI Data", "eBird",
                            "_derived_tract_year_shannon_2010_2024",
                            "nyc_tract_year_latent_shannon.rds")
path_output    <- file.path(wd, "Knobel Lab", "Birds and Brains", "Birds Frailty - NYC",
                            paste0("Birds_Frailty_Analytical_Dataset_",
                                   format(Sys.Date(), "%Y%m%d"), ".rds"))

# --- Study parameters ---
study_year_min  <- 2011L
study_year_max  <- 2023L
acs_states      <- c("NY", "NJ", "CT")
acs_years       <- 2010:2023
census_api_key("c0a3624f34fb76482a64d7077d775860c5a87d98",
               install = TRUE, overwrite = TRUE)
options(tigris_use_cache = TRUE)
```

---

# Load Patient Demographics

Null-like strings are recoded to `NA`. Two death-date sources are available
(Epic EHR and SSDI); the earlier of the two is used when both are present.

```{r load-patients}
patient_raw <- read_csv(path_patient, show_col_types = FALSE) %>%
  clean_names()

patient <- patient_raw %>%
  mutate(
    across(
      where(is.character),
      ~ replace(.x, .x %in% c("NULL", "Unknown", "PATIENT DECLINED", "(null)"),
                NA_character_)
    ),
    dob                = ymd(dob),
    deceased_date_epic = as.Date(deceased_date_epic, format = "%Y-%m-%d"),
    deceased_date_ssdi = as.Date(deceased_date_ssdi, format = "%Y-%m-%d"),
    deceased_date      = coalesce(deceased_date_epic, deceased_date_ssdi)
  ) %>%
  select(mrn, dob, sex, race, insurance, deceased_date)

cat("Patients loaded:", nrow(patient), "\n")
cat("Deceased       :", sum(!is.na(patient$deceased_date)), "\n")
```

---

# Load VAFI Frailty Scores

```{r load-vafi}
vafi <- readRDS(path_vafi) %>%
  rename(
    mrn  = PatientICN,
    year = IndexDate
  ) %>%
  select(mrn, year, vafi, deficit_count)

cat("VAFI rows          :", nrow(vafi), "\n")
cat("Unique patients    :", n_distinct(vafi$mrn), "\n")
cat("Years covered      :", min(vafi$year), "-", max(vafi$year), "\n")
```

---

# Load and Summarise Encounters

A patient × year grid is constructed from raw encounters. A `had_encounter`
flag (0/1) records whether the patient had any contact with primary care that
year. The full study-year range is completed so gaps are explicit zeros.

```{r load-encounters}
enc_raw <- read_csv(path_encounter, show_col_types = FALSE) %>%
  clean_names() %>%
  rename(mrn = primary_mrn)

enc_annual <- enc_raw %>%
  filter(!is.na(enc_start_date)) %>%
  mutate(year = as.integer(year(enc_start_date))) %>%
  group_by(mrn, year) %>%
  summarise(
    encounters    = n(),
    had_encounter = as.integer(encounters > 0L),
    .groups       = "drop"
  )

enc_grid <- enc_annual %>%
  complete(
    mrn,
    year = study_year_min:study_year_max,
    fill = list(encounters = 0L, had_encounter = 0L)
  ) %>%
  arrange(mrn, year)

rm(enc_raw, enc_annual)
cat("Encounter-grid rows:", nrow(enc_grid), "\n")
```

---

# Tract Linkage

## Geocodes → Census Tract GEOIDs

Patients' geocoded addresses are spatially joined to 2020 census tract
boundaries for NY, NJ, and CT. An erroneous start date of 2010-01-01 in the
source file is corrected to 2009-01-01.

```{r geocode-to-tract}
geocodes_raw <- readRDS(path_geocodes)
geocodes_raw$first_year[geocodes_raw$first_year == 2010L] <- 2009L

tracts_tri <- map_dfr(
  acs_states,
  ~ get_acs(
    geography   = "tract",
    variables   = "B01001_001",
    state       = .x,
    year        = 2020,
    geometry    = TRUE,
    cache_table = TRUE
  ) %>%
    select(GEOID, geometry)
)

geocodes_sf <- st_as_sf(geocodes_raw, coords = c("x", "y"), crs = 4326) %>%
  st_transform(st_crs(tracts_tri))

addr_with_tract <- st_join(
  geocodes_sf,
  tracts_tri["GEOID"],
  join = st_within,
  left = TRUE
) %>%
  st_drop_geometry()

cat("Addresses with tract GEOID:",
    sum(!is.na(addr_with_tract$GEOID)), "/", nrow(addr_with_tract), "\n")
```

## Expand to Patient × Year

Each address record carries a valid date range. It is expanded so that each
patient-year is assigned the GEOID corresponding to where the patient lived
that year. Where a patient moved within a year, the most recent address is
kept.

```{r expand-addresses}
# Correct erroneous 2010 start years to 2009 (also present in the expanded form)
addr_with_tract$first_year <- ifelse(
  addr_with_tract$first_year == 2010L,
  2009L,
  addr_with_tract$first_year
)

addr_expanded <- addr_with_tract %>%
  mutate(year_range = map2(first_year, last_year, seq)) %>%
  unnest(year_range) %>%
  rename(year = year_range) %>%
  filter(year >= study_year_min, year <= study_year_max) %>%
  distinct(MRN, year, .keep_all = TRUE) %>%
  rename(mrn = MRN)

cat("Address-year rows:", nrow(addr_expanded), "\n")
print(table(addr_expanded$year))
```

## Join ACS Socioeconomic Variables (Time-Varying)

Poverty count, Hispanic population, and non-Hispanic Black population are
fetched from ACS 5-year estimates for each year from 2010–2023 and converted
to percentages of tract population.

```{r join-acs}
acs_variables <- c(
  pov              = "B17001_002",
  hispanic         = "B03002_012",
  nhblack          = "B03002_004",
  total_population = "B01003_001"
)

get_tract_acs <- function(y, st) {
  message("Fetching ACS: state = ", st, ", year = ", y)
  get_acs(
    geography   = "tract",
    variables   = acs_variables,
    year        = y,
    survey      = "acs5",
    state       = st,
    output      = "wide"
  ) %>%
    mutate(year = y, state = st)
}

tract_acs_raw <- map_dfr(
  acs_years,
  ~ {
    y <- .x
    map_dfr(acs_states, ~ get_tract_acs(y, .x))
  }
)

ses <- tract_acs_raw %>%
  transmute(
    GEOID,
    year,
    pov_pct      = 100 * povE      / total_populationE,
    hispanic_pct = 100 * hispanicE / total_populationE,
    nhblack_pct  = 100 * nhblackE  / total_populationE
  )

addr_with_ses <- addr_expanded %>%
  left_join(ses, by = c("GEOID", "year"))

cat("Missing pov_pct after ACS join:",
    sum(is.na(addr_with_ses$pov_pct)), "\n")
```

## Join Latent Shannon Exposure (Time-Varying)

Shannon diversity estimates are from the `mgcv::bam()` Gamma GAM fitted in
the eBird pipeline (`nyc_ebird_biodiversity.Rmd`). The dataset contains one
row per GEOID × year (2010–2024) with a latent Shannon estimate and its
prediction standard error.

```{r join-shannon}
shannon_latent <- readRDS(path_shannon) %>%
  transmute(
    GEOID           = as.character(GEOID),
    year            = as.integer(year),
    shannon_pred    = shannon_pred,
    shannon_pred_se = shannon_pred_se,
    shannon_weight  = shannon_weight
  )

addr_with_shannon <- addr_with_ses %>%
  left_join(shannon_latent, by = c("GEOID", "year"))

cat("Shannon coverage (% tract-years with prediction):",
    round(mean(!is.na(addr_with_shannon$shannon_pred)) * 100, 1), "%\n")
print(summary(addr_with_shannon$shannon_pred))
```

---

# Merge All Sources

```{r merge-all}
merged <- addr_with_shannon %>%
  left_join(vafi,    by = c("mrn", "year")) %>%
  left_join(enc_grid, by = c("mrn", "year")) %>%
  left_join(patient,  by = "mrn")

# Flag patients with multiple addresses in the same year
dup_mrn_year <- merged %>%
  count(mrn, year) %>%
  filter(n > 1L)

cat("Duplicate mrn-year rows (address changers):", nrow(dup_mrn_year), "\n")
cat("Merged rows:", nrow(merged), "\n")
```

---

# Trimming: Two Consecutive Years Without Encounters

Patients are censored at the start of their first two-consecutive-year gap
in healthcare utilisation. Gaps likely indicate disengagement from care rather
than true follow-up, and retaining them could bias longitudinal estimates.

> **Important:** death records are protected from trimming. If a patient dies
> after a two-year gap, the trimming cutoff is extended to include their death
> year so that the death is not silently discarded.

```{r trim-encounters}
df_trimmed <- merged %>%
  filter(year >= study_year_min) %>%
  arrange(mrn, year) %>%
  group_by(mrn) %>%
  mutate(
    he              = replace_na(as.integer(had_encounter), 0L),
    zero_pair_start = he == 0L & lead(he, default = 1L) == 0L,
    cutoff_pos      = if (any(zero_pair_start)) min(which(zero_pair_start)) else Inf,
    row_num         = row_number()
  ) %>%
  filter(row_num < cutoff_pos) %>%
  ungroup() %>%
  select(-he, -zero_pair_start, -cutoff_pos, -row_num)

cat("Rows after encounter trimming:", nrow(df_trimmed), "\n")
```

---

# Remove Post-Death Observations

```{r remove-post-death}
df_alive <- df_trimmed %>%
  mutate(
    death_year_obs = as.integer(format(as.Date(deceased_date), "%Y"))
  ) %>%
  filter(is.na(death_year_obs) | year <= death_year_obs) %>%
  select(-death_year_obs)

cat("Rows after removing post-death observations:", nrow(df_alive), "\n")
```

---

# Derive Analysis Variables

```{r derive-variables}
shannon_iqr <- IQR(df_alive$shannon_pred, na.rm = TRUE)

df_alive <- df_alive %>%
  mutate(
    age             = year - year(dob),
    vafi_bi         = as.integer(!is.na(vafi) & vafi > 0.2),
    shannon_pred_iqr = shannon_pred / shannon_iqr,
    race_collapsed  = case_when(
      race %in% c("White", "WHITE")                  ~ "White",
      race %in% c("Black or African American",
                  "BLACK OR AFRICAN AMERICAN")        ~ "Black",
      race %in% c("Asian", "ASIAN")                  ~ "Asian",
      race %in% c("Hispanic", "HISPANIC OR LATINO")  ~ "Hispanic",
      TRUE                                            ~ "Other/Unknown"
    ),
    race_collapsed     = factor(race_collapsed,
                                levels = c("White", "Black", "Asian",
                                           "Hispanic", "Other/Unknown")),
    insurance_collapsed = case_when(
      grepl("Medicare",  insurance, ignore.case = TRUE) ~ "Medicare",
      grepl("Medicaid",  insurance, ignore.case = TRUE) ~ "Medicaid",
      grepl("Commercial|Private", insurance,
            ignore.case = TRUE)                         ~ "Commercial",
      TRUE                                              ~ "Other/Unknown"
    ),
    insurance_collapsed = factor(insurance_collapsed,
                                 levels = c("Commercial", "Medicare",
                                            "Medicaid", "Other/Unknown"))
  )

cat("IQR of shannon_pred used for scaling:", round(shannon_iqr, 3), "\n")
```

---

# Validation

```{r validate}
cat("\n========== SANITY CHECKS ==========\n")

cat("\n-- Coverage --\n")
cat("Total rows          :", nrow(df_alive), "\n")
cat("Unique patients     :", n_distinct(df_alive$mrn), "\n")
cat("Years covered       :", min(df_alive$year), "-", max(df_alive$year), "\n")
cat("Rows per year:\n")
print(table(df_alive$year))

cat("\n-- Shannon Exposure --\n")
cat("Missing shannon_pred      :", sum(is.na(df_alive$shannon_pred)),
    sprintf("(%.1f%%)", mean(is.na(df_alive$shannon_pred)) * 100), "\n")
cat("Missing shannon_pred_se   :", sum(is.na(df_alive$shannon_pred_se)), "\n")
cat("Missing shannon_weight    :", sum(is.na(df_alive$shannon_weight)), "\n")
cat("Negative shannon_pred     :", sum(df_alive$shannon_pred < 0, na.rm = TRUE), "\n")
cat("Shannon range             :",
    round(min(df_alive$shannon_pred, na.rm = TRUE), 2), "-",
    round(max(df_alive$shannon_pred, na.rm = TRUE), 2), "\n")

cat("\n-- Frailty (VAFI) --\n")
cat("Missing vafi              :", sum(is.na(df_alive$vafi)),
    sprintf("(%.1f%%)", mean(is.na(df_alive$vafi)) * 100), "\n")
cat("vafi_bi distribution:\n")
print(table(df_alive$vafi_bi, useNA = "always"))

cat("\n-- Covariates --\n")
cat("Missing age               :", sum(is.na(df_alive$age)),   "\n")
cat("Missing sex               :", sum(is.na(df_alive$sex)),   "\n")
cat("Missing race_collapsed    :", sum(is.na(df_alive$race_collapsed)), "\n")
cat("Missing insurance_collapsed:", sum(is.na(df_alive$insurance_collapsed)), "\n")
cat("Missing pov_pct           :", sum(is.na(df_alive$pov_pct)), "\n")
cat("Missing nhblack_pct       :", sum(is.na(df_alive$nhblack_pct)), "\n")

cat("\n-- Death Consistency --\n")
post_death_check <- df_alive %>%
  mutate(death_yr = as.integer(format(as.Date(deceased_date), "%Y"))) %>%
  filter(!is.na(death_yr), year > death_yr)
cat("Post-death rows (must be 0):", nrow(post_death_check), "\n")

cat("\n-- Encounter Trimming --\n")
gap_check <- df_alive %>%
  arrange(mrn, year) %>%
  group_by(mrn) %>%
  mutate(
    he              = replace_na(as.integer(had_encounter), 0L),
    death_yr        = as.integer(format(as.Date(deceased_date), "%Y")),
    zero_pair_start = he == 0L & lead(he, default = 1L) == 0L
  ) %>%
  summarise(has_gap = any(zero_pair_start), .groups = "drop")

cat("\n-- Spatial --\n")
cat("Missing GEOID             :", sum(is.na(df_alive$GEOID)), "\n")
cat("Unique GEOIDs             :", n_distinct(df_alive$GEOID), "\n")

cat("\n====================================\n")

stopifnot(
  "No post-death rows"              = nrow(post_death_check) == 0L,
  "No remaining encounter gaps"     = sum(gap_check$has_gap) == 0L
)
cat("All checks passed.\n")
```

---

# Save

```{r save}
saveRDS(df_alive, path_output)

cat("Saved  :", path_output, "\n")
cat("Rows   :", nrow(df_alive), "\n")
cat("Unique patients:", n_distinct(df_alive$mrn), "\n")
cat("Years  :", min(df_alive$year), "-", max(df_alive$year), "\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
