---
title: "3 — Bird Biodiversity and Frailty Severity: Negative Binomial GAM"
author: "Pablo Knobel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: united
---

# Summary

This script estimates the longitudinal association between residential bird
biodiversity (latent Shannon diversity index, IQR-scaled) and frailty severity
(VAFI deficit count, 0–30) in a repeated-measures cohort of older adults
receiving primary care at the Mount Sinai Health System.

The primary model is a negative binomial GAM (`mgcv::bam`, `nb()` family)
with a patient-level random intercept (`s(mrn, bs = "re")`), a penalised
temporal smooth (`s(year)`), and time-varying Shannon as the exposure.
Coefficients are exponentiated and reported as count ratios (CR); no offset
is used, so results represent multiplicative changes in the absolute deficit
count rather than a rate.

Three sensitivity analyses are run:

1. **Inverse-variance weighted** — downweights tract-years with high Shannon
   prediction uncertainty (`shannon_weight = 1 / SE²`)
2. **Restricted to ≥ 2 VAFI observations** — excludes patients observed only
   once, who contribute no within-person information
3. **3-year rolling mean Shannon** — replaces the annual estimate with a
   3-year trailing mean to reduce year-to-year exposure noise

Subgroup analyses stratify by age group, sex, race/ethnicity, and insurance.

---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(mgcv)
library(gratia)
library(tibble)
library(purrr)
library(stringr)
library(zoo)
```

```{r parameters}
# --- Data path ---
# --- Paths ---
wd      <- "C:/Users/knobep01/OneDrive - The Mount Sinai Hospital"
out_dir <- file.path(wd, "Knobel Lab", "Birds and Brains", "Birds Frailty - NYC", "outputs")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Automatically load the most recent analytical dataset
data_dir  <- file.path(wd, "Knobel Lab", "Birds and Brains", "Birds Frailty - NYC")
path_data <- tail(sort(list.files(
  data_dir,
  pattern = "^Birds_Frailty_Analytical_Dataset_.*\\.rds$",
  full.names = TRUE
)), 1)

cat("Loading:", path_data, "\n")

# --- Model parameters ---
year_smooth_k    <- 7L    # Basis dimension for temporal smooth
total_deficits   <- 30L   # Number of VAFI deficit items
rolling_window   <- 3L    # Years for rolling-mean sensitivity analysis
min_vafi_obs     <- 2L    # Minimum VAFI observations for restriction SA

# --- Subgroup definitions ---
# Each element: list(variable, label, reference level for factor)
subgroup_specs <- list(
  list(var = "age_group",           label = "Age group"),
  list(var = "sex",                 label = "Sex"),
  list(var = "race_collapsed",      label = "Race/ethnicity"),
  list(var = "insurance_collapsed", label = "Insurance")
)
```

---

# Load and Prepare Data

```{r load-data}
df_raw <- readRDS(path_data)

cat("Rows loaded     :", nrow(df_raw), "\n")
cat("Unique patients :", n_distinct(df_raw$mrn), "\n")
cat("Years covered   :", min(df_raw$year), "-", max(df_raw$year), "\n")
```
```{r}

# Create a clean numeric patient ID for random effects
# Avoids issues with non-numeric MRNs (e.g. "A" prefix)
mrn_lookup <- df_raw %>%
  distinct(mrn) %>%
  mutate(patient_id = row_number())

df_raw <- df_raw %>%
  left_join(mrn_lookup, by = "mrn")

cat("Unique patients    :", n_distinct(df_raw$patient_id), "\n")
cat("Any NA patient_id  :", sum(is.na(df_raw$patient_id)), "\n")



```

## Prepare Model Dataset

Complete-case restriction on the variables required for all models. The
patient identifier `mrn` is converted to a factor, which is required by
`mgcv` for the `bs = "re"` smooth.

```{r prepare-model-data}
df_model <- df_raw %>%
  filter(
    !is.na(deficit_count),
    !is.na(shannon_pred_iqr),
    !is.na(age),
    !is.na(sex),
    !is.na(race_collapsed),
    !is.na(insurance_collapsed),
    !is.na(pov_pct),
    !is.na(nhblack_pct)
  ) %>%
  mutate(
    mrn       = factor(mrn),
    age_group = cut(age,
                    breaks = c(-Inf, 64, 74, Inf),
                    labels = c("< 65", "65–74", "≥ 75"),
                    right  = TRUE)
  ) %>%
  arrange(mrn, year)

cat("Model rows      :", nrow(df_model), "\n")
cat("Unique patients :", n_distinct(df_model$mrn), "\n")

cat("\n-- Overdispersion check --\n")
cat("Mean deficit count :", round(mean(df_model$deficit_count), 2), "\n")
cat("Var deficit count  :", round(var(df_model$deficit_count),  2), "\n")
cat("Var/Mean ratio     :",
    round(var(df_model$deficit_count) / mean(df_model$deficit_count), 2),
    "(> 1 supports negative binomial over Poisson)\n")
```

## Add 3-Year Rolling Mean Shannon (Sensitivity Analysis 3)

The rolling mean is computed within patient using a 3-year trailing window
(`align = "right"`). Patient-years with fewer than 3 preceding observations
are assigned `NA` and excluded from that sensitivity analysis.

```{r rolling-mean-shannon}
df_model <- df_model %>%
  group_by(mrn) %>%
  mutate(
    shannon_3yr_mean = rollmeanr(shannon_pred_iqr, k = rolling_window,
                                 fill = NA, na.rm = TRUE)
  ) %>%
  ungroup()

cat("Patient-years with 3-year rolling mean:",
    sum(!is.na(df_model$shannon_3yr_mean)), "\n")
```

---

# Primary Model

## Model Formula

Time-varying Shannon enters as a linear parametric term. The temporal trend
is captured by a penalised cubic regression spline, and patient clustering
is handled by a random intercept smooth. No offset is used — results are
therefore count ratios (CR), not incidence rate ratios.

```{r primary-formula}
formula_primary <- deficit_count ~
  shannon_pred_iqr +
  s(year, bs = "ps", k = year_smooth_k) +
  age +
  sex +
  race_collapsed +
  insurance_collapsed +
  pov_pct +
  nhblack_pct +
  s(patient_id, bs = "re")
  
```

## Fit

`bam()` with `fREML` and `discrete = TRUE` is used for computational
efficiency with the large patient random effect.

```{r fit-primary, cache=TRUE}

m_primary <- bam(
  formula_primary,
  family   = nb(),
  data     = df_model,
  method   = "fREML",
  discrete = TRUE,
  cluster  = cl
)

summary(m_primary)
```

## Extract Shannon Coefficient

Coefficients are on the log scale; `exp()` converts to count ratios.
A CR < 1 indicates fewer deficits (lower frailty severity) per 1-IQR
increment in Shannon diversity.

```{r extract-primary}
extract_shannon_cr <- function(model, label) {
  s         <- summary(model)
  coef_row  <- s$p.table["shannon_pred_iqr", ]
  estimate  <- coef_row["Estimate"]
  se        <- coef_row["Std. Error"]
  tibble(
    model      = label,
    cr         = exp(estimate),
    ci_low     = exp(estimate - 1.96 * se),
    ci_high    = exp(estimate + 1.96 * se),
    pct_change = (exp(estimate) - 1) * 100,
    p_value    = coef_row["Pr(>|t|)"]
  )
}

results_primary <- extract_shannon_cr(m_primary, "Primary (time-varying Shannon)")
print(results_primary)
```

---

# Dose–Response Curve

A partial effect plot for Shannon diversity on the response scale, holding
all other covariates at their observed values (integrated marginal effect via
`gratia::draw()`).

```{r plot-shannon-smooth, fig.width = 7, fig.height = 4}
# SA4: Nonlinear Shannon dose-response (for visualization only)
m_sa4_nonlinear <- bam(
  deficit_count ~
    s(shannon_pred_iqr, bs = "tp", k = 5) +
    s(year, bs = "ps", k = year_smooth_k) +
    age +
    sex +
    race_collapsed +
    insurance_collapsed +
    pov_pct +
    nhblack_pct +
    s(mrn, bs = "re"),
  family   = nb(),
  data     = df_model,
  method   = "fREML",
  discrete = TRUE
)

fig_shannon_smooth <- draw(m_sa4_nonlinear, select = "s(shannon_pred_iqr)") +
  labs(
    title    = "Dose-response: Shannon diversity and VAFI deficit count",
    subtitle = "Sensitivity analysis — nonlinear smooth; shaded = 95% CI",
    x        = "Shannon diversity (IQR-scaled)",
    y        = "Partial effect (log scale)"
  ) +
  theme_minimal()

fig_shannon_smooth

ggsave(
  filename = file.path(out_dir, "fig_shannon_nonlinear.tiff"),
  plot     = fig_shannon_smooth,
  width    = 7, height = 4, dpi = 300, device = "tiff", bg = "white"
)
```

## Temporal Trend

```{r plot-year-smooth, fig.width = 7, fig.height = 4}
fig_year_smooth <- draw(m_primary, select = "s(year)") +
  labs(
    title    = "Temporal trend in VAFI deficit count",
    subtitle = "Penalised cubic regression spline; shaded = 95% CI",
    x        = "Year",
    y        = "Partial effect (log scale)"
  ) +
  theme_minimal()

fig_year_smooth

ggsave(
  filename = file.path(out_dir, "fig_year_trend.tiff"),
  plot     = fig_year_smooth,
  width    = 7, height = 4, dpi = 300, device = "tiff", bg = "white"
)
```

---

# Sensitivity Analyses

## SA 1 — Inverse-Variance Weighted

Downweights tract-years where the Shannon GAM prediction had high uncertainty.
`discrete = FALSE` is required when user-supplied weights are passed to `bam()`.

```{r sa1-weighted, cache=TRUE}
m_sa1_weighted <- gam(
  formula_primary,
  family   = nb(),
  weights  = shannon_weight,
  data     = df_model,
  method   = "fREML",
  discrete = FALSE
)

results_sa1 <- extract_shannon_cr(m_sa1_weighted,
                                   "SA1: Inverse-variance weighted")
print(results_sa1)
```

## SA 2 — Restricted to ≥ 2 VAFI Observations

Patients observed only once cannot contribute within-person variation to the
random intercept. Restricting to ≥ 2 observations confirms that results are
not driven by single-observation patients.

```{r sa2-restricted, cache=TRUE}
mrn_multi_obs <- df_model %>%
  group_by(mrn) %>%
  summarise(n_obs = n(), .groups = "drop") %>%
  filter(n_obs >= min_vafi_obs) %>%
  pull(mrn)

df_sa2 <- df_model %>%
  filter(mrn %in% mrn_multi_obs) %>%
  mutate(mrn = droplevels(mrn))

cat("Patients in SA2 (≥", min_vafi_obs, "obs):", n_distinct(df_sa2$mrn), "\n")
cat("Rows in SA2                           :", nrow(df_sa2), "\n")

m_sa2_restricted <- bam(
  formula_primary,
  family   = nb(),
  data     = df_sa2,
  method   = "fREML",
  discrete = TRUE,
  nthreads = max(1L, parallel::detectCores() - 1L)
)

results_sa2 <- extract_shannon_cr(m_sa2_restricted,
                                   paste0("SA2: Restricted to ≥",
                                          min_vafi_obs, " VAFI obs"))
print(results_sa2)
```

## SA 3 — 3-Year Rolling Mean Shannon

Replaces the annual Shannon estimate with a 3-year trailing mean to reduce
year-to-year measurement noise in the exposure.

```{r sa3-rolling, cache=TRUE}
formula_sa3 <- update(formula_primary, . ~ . - shannon_pred_iqr + shannon_3yr_mean)

df_sa3 <- df_model %>%
  filter(!is.na(shannon_3yr_mean)) %>%
  mutate(mrn = droplevels(mrn))

cat("Rows in SA3 (3-yr rolling mean available):", nrow(df_sa3), "\n")

m_sa3_rolling <- gam(
  formula_sa3,
  family   = nb(),
  data     = df_sa3,
  method   = "fREML",
  discrete = TRUE,
  nthreads = max(1L, parallel::detectCores() - 1L)
)

# Extract using shannon_3yr_mean coefficient
extract_rolling_cr <- function(model, label) {
  s        <- summary(model)
  coef_row <- s$p.table["shannon_3yr_mean", ]
  estimate <- coef_row["Estimate"]
  se       <- coef_row["Std. Error"]
  tibble(
    model      = label,
    cr         = exp(estimate),
    ci_low     = exp(estimate - 1.96 * se),
    ci_high    = exp(estimate + 1.96 * se),
    pct_change = (exp(estimate) - 1) * 100,
    p_value    = coef_row["Pr(>|t|)"]
  )
}

results_sa3 <- extract_rolling_cr(m_sa3_rolling, "SA3: 3-year rolling mean Shannon")
print(results_sa3)
```

## Sensitivity Results Table

```{r sa-results-table}
results_all <- bind_rows(
  results_primary,
  results_sa1,
  results_sa2,
  results_sa3
) %>%
  mutate(
    across(c(cr, ci_low, ci_high), ~ round(.x, 3)),
    pct_change = round(pct_change, 1),
    p_value    = round(p_value, 4),
    result     = sprintf("CR %.3f (%.3f–%.3f), p = %.4f",
                         cr, ci_low, ci_high, p_value)
  )

print(results_all %>% select(model, result, pct_change))
```

---

# Subgroup Analyses

Each subgroup is fit separately using the same primary model formula, but
with covariates that become single-level within a stratum removed
automatically. Results are pooled into a forest plot.

```{r subgroup-function}
# Fit the primary model within one stratum, dropping any covariate that has
# only one unique non-NA level (which would cause rank deficiency).
fit_stratum <- function(data, stratum_var, stratum_level) {

  df_sub <- data %>%
    filter(.data[[stratum_var]] == stratum_level) %>%
    mutate(mrn = droplevels(mrn))

  if (nrow(df_sub) < 50L || n_distinct(df_sub$mrn) < 10L) {
    return(NULL)   # too few observations to fit reliably
  }

  # Remove the stratification variable and any single-level covariates
  single_level_vars <- c(stratum_var,
    names(which(sapply(
      df_sub %>% select(sex, race_collapsed, insurance_collapsed, age_group),
      function(x) length(unique(na.omit(x))) < 2L
    )))
  )

  rhs_terms <- labels(terms(formula_primary))
  rhs_clean <- rhs_terms[!str_detect(rhs_terms,
                          paste(single_level_vars, collapse = "|"))]
  formula_sub <- as.formula(
    paste("deficit_count ~", paste(rhs_clean, collapse = " + "))
  )

  tryCatch({
    m <- bam(
      formula_sub,
      family   = nb(),
      data     = df_sub,
      method   = "fREML",
      discrete = TRUE,
      nthreads = max(1L, parallel::detectCores() - 1L)
    )
    extract_shannon_cr(m, paste0(stratum_var, ": ", stratum_level)) %>%
      mutate(
        subgroup_var   = stratum_var,
        subgroup_level = stratum_level,
        n_rows         = nrow(df_sub),
        n_patients     = n_distinct(df_sub$mrn)
      )
  }, error = function(e) {
    message("Model failed for ", stratum_var, " = ", stratum_level, ": ", e$message)
    NULL
  })
}
```

```{r run-subgroups, cache=TRUE}
subgroup_vars <- c("age_group", "sex", "race_collapsed", "insurance_collapsed")

subgroup_results <- map_dfr(
  subgroup_vars,
  function(var) {
    levels_var <- unique(na.omit(df_model[[var]]))
    map_dfr(levels_var, ~ fit_stratum(df_model, var, as.character(.x)))
  }
)

print(subgroup_results %>%
  select(subgroup_var, subgroup_level, n_patients, cr, ci_low, ci_high, p_value))
```

## Subgroup Forest Plot

```{r forest-plot, fig.width = 8, fig.height = 6}
subgroup_labels <- tibble(
  subgroup_var = c("age_group", "sex", "race_collapsed", "insurance_collapsed"),
  var_label    = c("Age group", "Sex", "Race/ethnicity", "Insurance")
)

plot_data <- subgroup_results %>%
  left_join(subgroup_labels, by = "subgroup_var") %>%
  mutate(
    label      = paste0(var_label, ": ", subgroup_level),
    label      = factor(label, levels = rev(unique(label))),
    sig        = p_value < 0.05
  )

fig_forest <- ggplot(plot_data,
                     aes(x = pct_change, y = label, colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(
    aes(xmin = (ci_low  - 1) * 100,
        xmax = (ci_high - 1) * 100),
    height = 0.3
  ) +
  geom_point(size = 2.5) +
  scale_colour_manual(
    values = c("TRUE" = "#B22222", "FALSE" = "#333333"),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05"),
    name   = NULL
  ) +
  labs(
    title    = "Subgroup analysis: Shannon diversity and VAFI deficit count",
    subtitle = "Count ratio per 1-IQR increment in Shannon; error bars = 95% CI",
    x        = "% change in deficit count per 1-IQR Shannon",
    y        = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank()
  )

fig_forest

ggsave(
  filename = file.path(out_dir, "fig_subgroup_forest.tiff"),
  plot     = fig_forest,
  width    = 8, height = 6, dpi = 300, device = "tiff", bg = "white"
)
```

---

# Model Diagnostics

```{r diagnostics, fig.width = 8, fig.height = 6}
# gam.check produces 4 diagnostic plots and prints basis dimension adequacy
gam.check(m_primary)
```

Concurvity between smooth terms (analogous to multicollinearity):

```{r concurvity}
concurvity(m_primary, full = FALSE)
```

---

# Summary Table

```{r summary-table}
results_summary <- results_all %>%
  select(model, cr, ci_low, ci_high, pct_change, p_value)

cat("\n===== PRIMARY AND SENSITIVITY ANALYSIS RESULTS =====\n")
print(results_summary, n = Inf)
cat("\nInterpretation: CR < 1 = fewer deficits per 1-IQR higher Shannon.\n")
cat("pct_change = (CR - 1) × 100 = % difference in deficit count.\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
