---
title: "3 — Bird Biodiversity and Frailty Severity: Negative Binomial GAM"
author: "Pablo Knobel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    theme: united
---

# Summary

This script estimates the longitudinal association between residential bird
biodiversity (latent Shannon diversity index, IQR-scaled) and frailty severity
(VAFI deficit count, 0–31) in a repeated-measures cohort of older adults
receiving primary care at the Mount Sinai Health System.

The primary model is an inverse-variance weighted negative binomial GAM
(`mgcv::gam`, `nb()` family) with a numeric patient-level random intercept
(`s(patient_id, bs = "re")`), a penalised temporal smooth (`s(year)`), and
time-varying annual Shannon as a linear parametric term. Coefficients are
exponentiated and reported as count ratios (CR); no offset is used.

**Sensitivity analyses (all inverse-variance weighted):**

- SA1: 3-year rolling mean Shannon exposure
- SA2: Nonlinear Shannon dose-response (smooth term, no random effect)
- SA3: Restricted to 2015 onward (addresses suspicious pre-2015 data quality)

**Age analyses (both inverse-variance weighted):**

- Interaction model: `shannon_pred_iqr × age_group`
- Stratified models per age group (< 65, 65–74, ≥ 75)

---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(mgcv)
library(gratia)
library(tibble)
library(purrr)
library(stringr)
library(zoo)
library(tableone)
library(flextable)
library(officer)
library(fst)
library(here)
```

```{r parameters}
# --- Paths ---
wd      <- "C:/Users/knobep01/OneDrive - The Mount Sinai Hospital"
out_dir <- file.path(wd, "Knobel Lab", "Birds and Brains",
                     "Birds Frailty - NYC", "outputs")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Automatically load the most recent analytical dataset
data_dir  <- file.path(wd, "Knobel Lab", "Birds and Brains", "Birds Frailty - NYC")
path_data <- tail(sort(list.files(
  data_dir,
  pattern    = "^Birds_Frailty_Analytical_Dataset_.*\\.rds$",
  full.names = TRUE
)), 1)

cat("Loading:", path_data, "\n")

# --- Mapping files (for deficit column names) ---
path_icd_map  <- here::here("VA-FI Mappings", "icd_code_vafi_mapping.fst")
path_prcd_map <- here::here("VA-FI Mappings", "procedure_code_vafi_mapping.fst")

cols_deficit <- union(
  fst::read.fst(path_icd_map)$Deficit,
  fst::read.fst(path_prcd_map)$Deficit
) %>% sort()

cat("Deficits in mapping files:", length(cols_deficit), "\n")

# --- Model parameters ---
year_smooth_k  <- 7L    # Basis dimension for temporal smooth
total_deficits <- 31L   # Number of VAFI deficit items
rolling_window <- 3L    # Years for rolling-mean sensitivity analysis
sa3_year_min   <- 2015L # Start year for SA3
```

---

# Load and Prepare Data

```{r load-data}
df_raw <- readRDS(path_data)

cat("Rows loaded     :", nrow(df_raw), "\n")
cat("Unique patients :", n_distinct(df_raw$mrn), "\n")
cat("Years covered   :", min(df_raw$year), "-", max(df_raw$year), "\n")
```

## Create Numeric Patient ID

MRNs include alphanumeric values (e.g. "A" prefix) that cannot be coerced to
numeric. A sequential `patient_id` is assigned to all patients to enable use
as a numeric random effect in `mgcv`.

```{r patient-id}
mrn_lookup <- df_raw %>%
  distinct(mrn) %>%
  mutate(patient_id = as.numeric(row_number()))

df_raw <- df_raw %>%
  left_join(mrn_lookup, by = "mrn")

cat("Unique patient_ids :", n_distinct(df_raw$patient_id), "\n")
cat("Any NA patient_id  :", sum(is.na(df_raw$patient_id)), "\n")
```

## Prepare Model Dataset

Complete-case restriction on all variables required across models.

```{r prepare-model-data}
df_model <- df_raw %>%
  filter(
    !is.na(deficit_count),
    !is.na(shannon_pred_iqr),
    !is.na(shannon_weight),
    !is.na(age),
    !is.na(sex),
    !is.na(race_collapsed),
    !is.na(insurance_collapsed),
    !is.na(pov_pct),
    !is.na(nhblack_pct),
    !is.na(patient_id)
  ) %>%
  mutate(
    age_group = cut(age,
                    breaks = c(-Inf, 64, 74, Inf),
                    labels = c("< 65", "65–74", "≥ 75"),
                    right  = TRUE)
  ) %>%
  arrange(patient_id, year)

cat("Model rows      :", nrow(df_model), "\n")
cat("Unique patients :", n_distinct(df_model$patient_id), "\n")

cat("\n-- Overdispersion check --\n")
cat("Mean deficit count :", round(mean(df_model$deficit_count), 2), "\n")
cat("Var deficit count  :", round(var(df_model$deficit_count),  2), "\n")
cat("Var/Mean ratio     :",
    round(var(df_model$deficit_count) / mean(df_model$deficit_count), 2),
    "(> 1 supports negative binomial)\n")
```

## Add 3-Year Rolling Mean Shannon

```{r rolling-mean-shannon}
df_model <- df_model %>%
  group_by(patient_id) %>%
  mutate(
    shannon_3yr_mean = rollmeanr(shannon_pred_iqr, k = rolling_window,
                                 fill = NA, na.rm = TRUE)
  ) %>%
  ungroup()

cat("Patient-years with 3-year rolling mean:",
    sum(!is.na(df_model$shannon_3yr_mean)), "\n")
```

---

# Table 1

Characteristics at cohort entry (first observed patient-year).

```{r table1}
df_table1 <- df_model %>%
  arrange(patient_id, year) %>%
  group_by(patient_id) %>%
  slice(1) %>%
  ungroup()

cat("Patients in Table 1:", nrow(df_table1), "\n")

tab1_vars <- c("age", "age_group", "sex", "race_collapsed",
               "insurance_collapsed", "pov_pct", "nhblack_pct",
               "shannon_pred_iqr", "deficit_count", "vafi")

tab1_cat <- c("age_group", "sex", "race_collapsed", "insurance_collapsed")

tab1 <- CreateTableOne(
  vars       = tab1_vars,
  data       = df_table1,
  factorVars = tab1_cat
)

tab1_df <- print(tab1, quote = FALSE, noSpaces = TRUE,
                 printToggle = FALSE) %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "Variable")

tab1_ft <- flextable(tab1_df) %>%
  autofit() %>%
  theme_booktabs()

doc <- read_docx() %>%
  body_add_par(
    "Table 1. Characteristics at cohort entry",
    style = "heading 1"
  ) %>%
  body_add_flextable(tab1_ft)

print(doc, target = file.path(out_dir, "table1.docx"))
cat("Table 1 saved.\n")
```

---

# Primary Model

Annual Shannon enters as a linear parametric term. Inverse-variance weights
downweight tract-years with high Shannon prediction uncertainty. `gam()` is
used rather than `bam()` as the latter does not support user-supplied weights
reliably.

```{r primary-formula}
formula_primary <- deficit_count ~
  shannon_pred_iqr +
  s(year, bs = "ps", k = year_smooth_k) +
  age +
  sex +
  race_collapsed +
  insurance_collapsed +
  pov_pct +
  nhblack_pct +
  s(patient_id, bs = "re")
```

```{r fit-primary, cache=TRUE}
m_primary <- gam(
  formula_primary,
  family  = nb(),
  weights = shannon_weight,
  data    = df_model,
  method  = "REML"
)

summary(m_primary)
```

## Extract Shannon Coefficient

```{r extract-shannon}
extract_shannon_cr <- function(model, label, coef_name = "shannon_pred_iqr") {
  s        <- summary(model)
  coef_row <- s$p.table[coef_name, ]
  estimate <- coef_row["Estimate"]
  se       <- coef_row["Std. Error"]
  tibble(
    model      = label,
    cr         = exp(estimate),
    ci_low     = exp(estimate - 1.96 * se),
    ci_high    = exp(estimate + 1.96 * se),
    pct_change = (exp(estimate) - 1) * 100,
    p_value    = coef_row["Pr(>|t|)"]
  )
}

results_primary <- extract_shannon_cr(m_primary, "Primary (annual Shannon, IVW)")
print(results_primary)
```

---

# Plots

## Shannon Effect (Coefficient Plot)

```{r plot-shannon-coef, fig.width = 7, fig.height = 3}
fig_shannon_coef <- ggplot(results_primary,
                           aes(x = pct_change, y = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(
    aes(xmin = (ci_low  - 1) * 100,
        xmax = (ci_high - 1) * 100),
    height = 0.2
  ) +
  geom_point(size = 3) +
  labs(
    title    = "Effect of Shannon diversity on VAFI deficit count",
    subtitle = "IVW negative binomial GAM; count ratio per 1-IQR Shannon increment",
    x        = "% change in deficit count per 1-IQR Shannon",
    y        = NULL
  ) +
  theme_minimal()

fig_shannon_coef

ggsave(
  filename = file.path(out_dir, "fig_shannon_effect.tiff"),
  plot     = fig_shannon_coef,
  width    = 7, height = 3, dpi = 300, device = "tiff", bg = "white"
)
```

## Temporal Trend

```{r plot-year-smooth, fig.width = 7, fig.height = 4}
fig_year_smooth <- draw(m_primary, select = "s(year)") +
  labs(
    title    = "Temporal trend in VAFI deficit count",
    subtitle = "Penalised cubic regression spline; shaded = 95% CI",
    x        = "Year",
    y        = "Partial effect (log scale)"
  ) +
  theme_minimal()

fig_year_smooth

ggsave(
  filename = file.path(out_dir, "fig_year_trend.tiff"),
  plot     = fig_year_smooth,
  width    = 7, height = 4, dpi = 300, device = "tiff", bg = "white"
)
```

---

# Sensitivity Analyses

All sensitivity analyses use inverse-variance weighting.

## SA1 — 3-Year Rolling Mean Shannon

Replaces the annual Shannon estimate with a 3-year trailing mean to reduce
year-to-year measurement noise.

```{r sa1-rolling, cache=TRUE}
formula_sa1 <- update(formula_primary,
                      . ~ . - shannon_pred_iqr + shannon_3yr_mean)

df_sa1 <- df_model %>%
  filter(!is.na(shannon_3yr_mean))

cat("Rows in SA1 (3-yr rolling mean available):", nrow(df_sa1), "\n")

m_sa1_rolling <- gam(
  formula_sa1,
  family  = nb(),
  weights = shannon_weight,
  data    = df_sa1,
  method  = "REML"
)

results_sa1 <- extract_shannon_cr(m_sa1_rolling,
                                   "SA1: 3-year rolling mean Shannon (IVW)",
                                   coef_name = "shannon_3yr_mean")
print(results_sa1)
```

## SA2 — Nonlinear Shannon Dose-Response

Shannon enters as a smooth term to assess non-linearity. The patient random
effect is dropped as it suppresses the between-person Shannon signal when
fitted jointly with the smooth. This model captures both within- and
between-person dose-response.

```{r sa2-nonlinear, cache=TRUE}
formula_sa2 <- deficit_count ~
  s(shannon_pred_iqr, bs = "tp", k = 5) +
  s(year, bs = "ps", k = year_smooth_k) +
  age +
  sex +
  race_collapsed +
  insurance_collapsed +
  pov_pct +
  nhblack_pct

m_sa2_nonlinear <- gam(
  formula_sa2,
  family  = nb(),
  weights = shannon_weight,
  data    = df_model,
  method  = "REML"
)

fig_shannon_smooth <- draw(m_sa2_nonlinear,
                           select = "s(shannon_pred_iqr)") +
  labs(
    title    = "Dose-response: Shannon diversity and VAFI deficit count",
    subtitle = "SA2 — nonlinear smooth, IVW; shaded = 95% CI",
    x        = "Shannon diversity (IQR-scaled)",
    y        = "Partial effect (log scale)"
  ) +
  theme_minimal()

fig_shannon_smooth

ggsave(
  filename = file.path(out_dir, "fig_shannon_nonlinear.tiff"),
  plot     = fig_shannon_smooth,
  width    = 7, height = 4, dpi = 300, device = "tiff", bg = "white"
)
```

## SA3 — Restricted to 2015 Onward

The temporal trend suggests potential data quality issues before 2015.
This analysis confirms robustness to exclusion of the early period.

```{r sa3-post2015, cache=TRUE}
df_sa3 <- df_model %>%
  filter(year >= sa3_year_min)

cat("Rows in SA3 (≥ 2015):", nrow(df_sa3), "\n")
cat("Unique patients      :", n_distinct(df_sa3$patient_id), "\n")

m_sa3_post2015 <- gam(
  formula_primary,
  family  = nb(),
  weights = shannon_weight,
  data    = df_sa3,
  method  = "REML"
)

results_sa3 <- extract_shannon_cr(m_sa3_post2015,
                                   "SA3: Restricted to 2015 onward (IVW)")
print(results_sa3)
```

## Sensitivity Results Table

```{r sa-results-table}
results_sa <- bind_rows(results_primary, results_sa1, results_sa3) %>%
  mutate(
    across(c(cr, ci_low, ci_high), ~ round(.x, 3)),
    pct_change = round(pct_change, 1),
    p_value    = round(p_value, 4),
    result     = sprintf("CR %.3f (%.3f–%.3f), p = %.4f",
                         cr, ci_low, ci_high, p_value)
  )

cat("\n===== SENSITIVITY ANALYSIS RESULTS =====\n")
print(results_sa %>% select(model, result, pct_change), n = Inf)
```

---

# Age Analyses

## Interaction Model

Tests whether the Shannon–frailty association differs by age group.
`age_group` replaces continuous `age` as a covariate to allow the
interaction term.

```{r age-interaction, cache=TRUE}
formula_interaction <- deficit_count ~
  shannon_pred_iqr * age_group +
  s(year, bs = "ps", k = year_smooth_k) +
  sex +
  race_collapsed +
  insurance_collapsed +
  pov_pct +
  nhblack_pct +
  s(patient_id, bs = "re")

m_interaction <- gam(
  formula_interaction,
  family  = nb(),
  weights = shannon_weight,
  data    = df_model,
  method  = "REML"
)

summary(m_interaction)

# Interaction p-value
anova_interaction <- anova(m_primary, m_interaction, test = "Chisq")
print(anova_interaction)
```


```{r}

# Compute Shannon CR for each age group from interaction model
# Reference group (<65): main Shannon coefficient
# Other groups: main + interaction term

coefs <- coef(m_interaction)
vcov  <- vcov(m_interaction)

age_groups <- c("< 65", "65–74", "≥ 75")

interaction_results <- map_dfr(age_groups, function(ag) {
  if (ag == "< 65") {
    est <- coefs["shannon_pred_iqr"]
    se  <- sqrt(vcov["shannon_pred_iqr", "shannon_pred_iqr"])
  } else {
    int_term <- paste0("shannon_pred_iqr:age_group", ag)
    est <- coefs["shannon_pred_iqr"] + coefs[int_term]
    se  <- sqrt(
      vcov["shannon_pred_iqr", "shannon_pred_iqr"] +
      vcov[int_term, int_term] +
      2 * vcov["shannon_pred_iqr", int_term]
    )
  }
  tibble(
    age_group  = ag,
    cr         = exp(est),
    ci_low     = exp(est - 1.96 * se),
    ci_high    = exp(est + 1.96 * se),
    pct_change = (exp(est) - 1) * 100,
    p_value    = 2 * pnorm(-abs(est / se))
  )
})

print(interaction_results)

```


```{r}
# Generate prediction grid
shannon_range <- seq(min(df_model$shannon_pred_iqr),
                     max(df_model$shannon_pred_iqr),
                     length.out = 100)

pred_grid <- expand_grid(
  shannon_pred_iqr    = shannon_range,
  age_group           = levels(df_model$age_group)
) %>%
  mutate(
    year                = median(df_model$year),
    sex                 = "Female",
    race_collapsed      = "White",
    insurance_collapsed = "Commercial",
    pov_pct             = median(df_model$pov_pct,     na.rm = TRUE),
    nhblack_pct         = median(df_model$nhblack_pct, na.rm = TRUE),
    patient_id          = 0   # population-level prediction (re = 0)
  )

pred_grid <- pred_grid %>%
  bind_cols(
    as.data.frame(predict(m_interaction, newdata = pred_grid,
                          se.fit = TRUE, exclude = "s(patient_id)"))
  ) %>%
  mutate(
    fit_exp  = exp(fit),
    lwr      = exp(fit - 1.96 * se.fit),
    upr      = exp(fit + 1.96 * se.fit)
  )

pred_grid <- pred_grid %>%
  mutate(age_group = factor(age_group, levels = c("< 65", "65–74", "≥ 75")))

fig_simple_slopes <- ggplot(
  pred_grid,
  aes(x = shannon_pred_iqr, y = fit_exp,
      colour = age_group, fill = age_group)
) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) +
  labs(
    title    = "Simple slopes: Shannon diversity and VAFI deficit count by age group",
    subtitle = "Predicted values at median covariates; shaded = 95% CI",
    x        = "Shannon diversity (IQR-scaled)",
    y        = "Predicted VAFI deficit count",
    colour   = "Age group",
    fill     = "Age group"
  ) +
  theme_minimal()

fig_simple_slopes

ggsave(
  filename = file.path(out_dir, "fig_simple_slopes.tiff"),
  plot     = fig_simple_slopes,
  width    = 7, height = 4, dpi = 300, device = "tiff", bg = "white"
)

```


---

# Model Diagnostics

```{r diagnostics, fig.width = 8, fig.height = 6}
gam.check(m_primary)
```

```{r concurvity}
concurvity(m_primary, full = FALSE)
```

---

# Session Info

```{r session-info}
sessionInfo()
```
